Pushdown Control-Flow Analysis via Refunctionalization (Functional Pearl)

### 2018-02-25 paper.tex
update

  We show their correspondence and how to syntactically transform a small-step 
  abstracting abstract machine to a big-step abstracting definitional 
  interpreter.
  The transformations include fusing, disentangling, refunctionalizing 
  and monadification(un-CPS), and all of them properly handle the 
  collecting semantics and the non-determinism of abstract interpretation. 
  After each transformation, we also obtain an intermediary form of 
  abstract interpreter.
  Remarkably, follow the idea that evaluation contexts are defuntionalized 
  continuations, we reveal how precise call/return match in control-flow 
  analysis is obtained by refunctionalizing a small-step abstracting 
  abstract machine,  % with unbounded stack
  as well as explain how it been lost by defunctionalizing the higher-order
  continuations into an algebraic data type. Further, we show how to turn 
  precise call/return match into computable analysis.

@@

the essence of abstracting defintional interpreter.

@@

\item We present an abstract interpreter obtained from refunctionalizing
  an small-step abstracting abstract machine. 
  This abstract interpreter naturally has pushdown control-flow property
  since we represent the evluation contexts of analyzed language using
  continuations of the meta-language, and therefore sequentialize the
  analyzing.
  We also analyze its termination, soundness and complexity.

@@

In this section, we frist describe the target language we will
analyze and a concrete CESK machine as operational semantics for it.
Then we review some recent works in control-flow analysis for
higher-order languages:
1) abstracting abstract machines (AAM) as an abstract interpretation
counterpart of CESK machines, and
2) the precise call/return match problem in control-flow analysis and its existing solutions.
We will show in section 3 that the abstracting abstract machines form the basis of our transformations,
and call/return match is obtained during transformation.

@@

Tranditionally, continuation-passing style (CPS) is a popular intermediate representation
for analyzing functional programs because of it exposes control transfer explicitly
and simplifies the analysis \cite{Shivers:1991:SSC:115865.115884, Shivers:1988:CFA:53990.54007}.
Here, we choose to use administrative normal form (ANF) which is a direct-style intermediate representation 
as target for clarity but without losing simplicty and generality.
Although we just show the core calculus language, but it can be easily extended
to support recursive bindings, conditionals, primitive types and operations on primitive types.
The transformations we will show in the rest of this paper
also work on abstract machines for plain, direct-style $\lambda$-calculus languages.
TODO: add letrec?

@@

\textit{C}ontrol \textit{E}nvionment \textit{S}tore \textit{K}ontinuation (CESK) machine is an
abstract machine for describing semantics of and evaluating $\lambda$-calculus \cite{felleisen1987calculus}.
CESK machine models program execution as state transitions in a small-step fashion. As its name suggested,
a machine state has four components: 1) Control is the expression currently being evaluated.
2) Environment is the a map that contains the address of a variable that in the lexical
scope of control.
3) Store models the heap of a program, and it is also a map but maps from addresses to values.
The addresses are infinite numbers starting from 0.
In our toy language, the only catagory of value is closure, i.e., a function paired with an environment.
4) Continuation represents the program stack.

@@

To run the program, we first use function \texttt{inject} to construct an initial machine
state given an expression \texttt{e}. The initial state contains an emtpy environment, 
store and stack.
\begin{verbatim}
def inject(e: Expr): State = State(e, Map(), Map(), Nil)
\end{verbatim}

@@

\subsection{Abstracting Abstract Machine} \label{aam}
Abstracting abstract machine (AAM) is a systematic methodology that derives sound 
abstract interpreters for higher-order functional languages from concrete 
abstract machines \cite{van2012systematic, van2010abstracting}. 
An abstracting abstract machine implements computable abstract semantics that 
approximates the runtime behaviors of programs.
Since the state space of concrete execution is possibily infinite,
so to analyze programs, the key insight of AAM approach is to allocate both bindings
and continuations on the store, and then bound the addresses space to be finite.
Thus the approximated machine-state space is finite, and computable.

@@

As we mentioned previously, the space of addresses is finite in AAM.
Binding addresses are parameterized by variable names and the time that creates this binding.
Continuation addresses \texttt{KAddr} has two variants: 
1) \texttt{Halt} which corresponds to the empty stack, and
2) \texttt{ContAddr} consists of the target expression of callee and the time that
creates this address. (TODO: why target expression, allocation polyvariance)

@@

To run (analyze) a program, as CESK machine, we use \texttt{inject} to construct
the initial state given a program. 
Note that the inital continuation store has a special mapping that maps continuation address
\texttt{Halt} to empty set of continuations.
We also provide an empty program contour as inital time.
\begin{verbatim}
def inject(e: Expr): State = 
  State(e, Map(), 
        Store[BAddr, Storable](Map()), 
        Store[KAddr, Cont](Map(Halt -> Set())), 
        Halt, List())
\end{verbatim}

@@

To see the problem, we can consider the following example:

\begin{verbatim}
(let ([x (id 1)])
  (let ([y (id 2)])
    x))
\end{verbatim}

return flow of \texttt{id} is polluted, and returns to both \texttt{x} and
\texttt{y}. TODO: explain more

@@

\begin{verbatim}
case class State(e: Expr, env: Env, bstore: BStore, konts: List[Frame], time: Time)

def step(s: State): List[State] = {
  val new_time = tick(s)
  s match {
    case State(Let(x, App(f, ae), e), env, bstore, konts, time) if isAtomic(ae) =>
      for (Clos(Lam(v, body), env_c) <- atomicEval(f, env, bstore).toList) yield {
        val frame = Frame(x, e, env)
        val baddr = allocBind(v, new_time)
        val new_env = env_c + (v -> baddr)
        val new_store = bstore.update(baddr, atomicEval(ae, env, bstore))
        State(body, new_env, new_store, frame::konts, new_time)
      }
    case State(ae, env, bstore, konts, time) if isAtomic(ae) =>
      konts match {
        case Nil => List()
        case Frame(x, e, env_f)::konts =>
          val baddr = allocBind(x, new_time)
          val new_env = env_f + (x -> baddr)
          val new_store = bstore.update(baddr, atomicEval(ae, env, bstore))
          List(State(e, new_env, new_store, konts, new_time))
      }
  }
}
\end{verbatim}

@@

We can transform the refunctionalized AAM to direct style, but explictly use 
side effects such as assignment and mutation to handle collecting semantics and non-determinism.

>>>>

  We show their correspondence and how to syntactically transform a small-step 
  abstracting abstract machine to a big-step abstracting definitional 
  interpreter.
  The transformations include fusing, disentangling, refunctionalizing 
  and monadification(un-CPS), and all of them properly handle the 
  collecting semantics and the non-determinism of abstract interpretation. 
  After each transformation, we also obtain an intermediary form of 
  abstract interpreter.
  Remarkably, follow the idea that evaluation contexts are defunctionalized 
  continuations, we reveal how precise call/return match in control-flow 
  analysis is obtained by refunctionalizing a small-step abstracting 
  abstract machine,  % with unbounded stack
  as well as explain how it been lost by defunctionalizing the higher-order
  continuations into an algebraic data type. Further, we show how to turn 
  precise call/return match into computable analysis.

@@

the essence of abstracting definitional interpreter.

@@

\item We present an abstract interpreter obtained from refunctionalizing
  an small-step abstracting abstract machine. 
  This abstract interpreter naturally has pushdown control-flow property
  since we represent the evaluation contexts of analyzed language using
  continuations of the meta-language, and therefore sequentialize the
  analyzing.
  We also analyze its termination, soundness and complexity.

@@

In this section, we first describe the target language we will
analyze and a concrete CESK machine as operational semantics for it.
Then we review some recent works in control-flow analysis for
higher-order languages:
1) abstracting abstract machines (AAM) as an abstract interpretation
counterpart of CESK machines, and
2) the precise call/return match problem in control-flow analysis and its existing solutions.
We will show in section 3 that the abstracting abstract machines form the basis of our transformations,
and call/return match is obtained during transformation.

@@

Traditionally, continuation-passing style (CPS) is a popular intermediate representation
for analyzing functional programs because of it exposes control transfer explicitly
and simplifies the analysis \cite{Shivers:1991:SSC:115865.115884, Shivers:1988:CFA:53990.54007}.
Here, we choose to use administrative normal form (ANF) which is a direct-style intermediate representation 
as target for clarity but without losing simplicity and generality.
Although we just show the core calculus language, but it can be easily extended
to support recursive bindings, conditionals, primitive types and operations on primitive types.
The transformations we will show in the rest of this paper
also work on abstract machines for plain, direct-style $\lambda$-calculus languages.
TODO: add letrec?

@@

\textit{C}ontrol \textit{E}nvionment \textit{S}tore \textit{K}ontinuation (CESK) machine is an
abstract machine for describing semantics of and evaluating $\lambda$-calculus \cite{felleisen1987calculus}.
CESK machine models program execution as state transitions in a small-step fashion. As its name suggested,
a machine state has four components: 1) Control is the expression currently being evaluated.
2) Environment is the a map that contains the address of a variable that in the lexical
scope of control.
3) Store models the heap of a program, and it is also a map but maps from addresses to values.
The addresses are infinite numbers starting from 0.
In our toy language, the only category of value is closure, i.e., a function paired with an environment.
4) Continuation represents the program stack.

@@

To run the program, we first use function \texttt{inject} to construct an initial machine
state given an expression \texttt{e}. The initial state contains an empty environment, 
store and stack.
\begin{verbatim}
def inject(e: Expr): State = State(e, Map(), Map(), Nil)
\end{verbatim}

@@

\subsection{Abstracting Abstract Machine} \label{aam}
Abstracting abstract machine (AAM) is a systematic methodology that derives sound 
abstract interpreters for higher-order functional languages from concrete 
abstract machines \cite{van2012systematic, van2010abstracting}. 
An abstracting abstract machine implements computable abstract semantics that 
approximates the runtime behaviors of programs.
Since the state space of concrete execution is possibly infinite,
so to analyze programs, the key insight of AAM approach is to allocate both bindings
and continuations on the store, and then bound the addresses space to be finite.
Thus the approximated machine-state space is finite, and computable.

@@

As we mentioned previously, the space of addresses is finite in AAM.
Binding addresses are parameterized by variable names and the time that creates this binding,
which are both finite.
Continuation addresses \texttt{KAddr} has two variants: 
1) \texttt{Halt} which corresponds to the empty stack, and
2) \texttt{ContAddr} consists of the entry expression of callee and the time that
creates this address, and they are also finite. 
(TODO: why target expression, allocation polyvariance)

@@

To run (analyze) a program, as CESK machine, we use \texttt{inject} to construct
the initial state given a program. 
Note that the initial continuation store has a special mapping that maps continuation address
\texttt{Halt} to empty set of continuations.
We also provide an empty program contour as initial time.
\begin{verbatim}
def inject(e: Expr): State = 
  State(e, Map(), 
        Store[BAddr, Storable](Map()), 
        Store[KAddr, Cont](Map(Halt -> Set())), 
        Halt, List())
\end{verbatim}

@@

To see the problem of spurious return-flows, we can consider the following example:

\begin{verbatim}
(let ([id (lambda (z) z)])
  (let ([x (id 1)])
    (let ([y (id 2)])
      x)))
\end{verbatim}

In traditional $k$-CFA algorithm and the AAM we showed in section \ref{aam}, 
the return-flow of invocation \texttt{(id 2)} 
is going to both call-sites \texttt{(id 1)} and \texttt{(id 2)}. Therefore, the
returned value \texttt{2} for variable \texttt{y} is also propagated to 
variable \texttt{x} and then imprecise analysis result is arose.
This return-flow merging is inevitable even we increase the context-sensitivity.
If we use the monovariant analysis 0CFA, the analyzing result would be \texttt{x} and 
\texttt{y} point to set \texttt{\{1, 2\}}, because it not distinguished 
\texttt{z} at different call-sites.
Under $1$-CFA, the algorithm is able to distinguish that variable \texttt{z} of function
\texttt{id} has two different values at two call-sites, so variable \texttt{y} 
would not be polluted by \texttt{1}, but it is still helplessness for distinguish
return-flows.
TODO: do need more details to explain this example? (as in P4F paper)

@@

A naive solution to is to use precise call stack modeling as we did in concrete
CESK machine. Instead of allocating continuations on store and recording an 
address of continuations, we

In the definition of state \texttt{State}, continuation store disappears, and
\texttt{konts} becomes a list of frames. Other components keep the same.

\begin{verbatim}
case class State(e: Expr, env: Env, bstore: BStore, konts: List[Frame], time: Time)
\end{verbatim}

the only non-determinism in our state transition function is when dereferencing
the callee \texttt{f} from store at a call-site.

\begin{verbatim}
def step(s: State): List[State] = {
  val new_time = tick(s)
  s match {
    case State(Let(x, App(f, ae), e), env, bstore, konts, time) if isAtomic(ae) =>
      for (Clos(Lam(v, body), env_c) <- atomicEval(f, env, bstore).toList) yield {
        val frame = Frame(x, e, env)
        val baddr = allocBind(v, new_time)
        val new_env = env_c + (v -> baddr)
        val new_store = bstore.update(baddr, atomicEval(ae, env, bstore))
        State(body, new_env, new_store, frame::konts, new_time)
      }
    case State(ae, env, bstore, konts, time) if isAtomic(ae) =>
      konts match {
        case Nil => List()
        case Frame(x, e, env_f)::konts =>
          val baddr = allocBind(x, new_time)
          val new_env = env_f + (x -> baddr)
          val new_store = bstore.update(baddr, atomicEval(ae, env, bstore))
          List(State(e, new_env, new_store, konts, new_time))
      }
  }
}
\end{verbatim}

@@

We can transform the refunctionalized AAM to direct style, but explicitly use 
side effects such as assignment and mutation to handle collecting semantics and non-determinism.
